A Streaming Algorithm for 2-Center with Outliers in High Dimensions

### 2015-05-02 paper.tex
New edits + page number shrink

@@

\newcommand{\textproc}{\textsc}
\newcommand{\Call}[2]{\textsc{#1}(#2)}
\renewcommand{\algorithmicforall}{{\bf for each}}

@@

\begin{abstract}
We study the Euclidean 2-center problem with outliers in the data stream model. 
Given a stream of points in arbitrary $d$ dimensions, the goal is to find two congruent balls 
of minimum radius covering all but $z$ points (\Cref{fig:definition}). 
We provide a $(1.8+\eps)$-approximation streaming algorithm for the problem, 
improving upon the previous $(4 + \eps)$-approximation algorithm available for the problem.
The space complexity and update time of our algorithm is $\poly(d, z, {1 \over \eps})$,
independent of the size of the stream.
\end{abstract}

@@

\section{Introduction}
The $k$-center problem---covering a set of points 
using $k$ congruent balls of minimum radius---is a fundamental problem,
arising in many applications 
such as data mining, unsupervised learning, statistics, and image processing.
In real-world applications, where input data is often noisy, 
it is very important to consider outliers, 
as even a small number of outliers can greatly affect the quality of the solution.
The $k$-center problem is particularly very sensitive to outliers,
and even a constant number of outliers can increase the radius of the $k$-center unboundedly.
Therefore, it is natural to consider the following generalization of
the the $k$-center problem: % which we call \emph{$k$-center with $z$ outliers}:
given a set $P$ of $n$ points in arbitrary $d$ dimensions, 
find $k$ congruent balls of minimum radius 
to cover at least $n - z$ points of $P$.
%One of $k$-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set.

In this paper, we focus on the \emph{data stream model}
where only a single pass over the input data is allowed,
and we have only a limited amount of working space available.
%which is typically sublinear in the input size.
This model is specially useful for processing large data sets,
as it does not need the entire data set to be stored in memory.

@@

The Euclidean $k$-center problem has been extensively studied in the literature.
%There is a considerable amount of work on the geometric $k$-center problem.
If $k$ is part of the input,
the problem in known to be NP-hard in two or more dimensions~\cite{fowler1981optimal},
and is even hard to approximate to within a factor close to 2, 
unless P$\,=\,$NP~\cite{feder1988optimal}.
Factor-2 approximation algorithms are available for the problem 
in any dimension~\cite{gonzalez1985clustering,feder1988optimal}.
%The $k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$~\cite{bern1996approximation}.
%Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, %there are more efficient algorithms.
better solutions are available.
The 1-center problem in fixed dimensions is known to be LP-type 
and can be solved in $O(n)$ time~\cite{chazelle1996linear}.
For 2-center in the plane, the current best algorithm
runs in $\O(n \log^2 n \log^2 \log n)$ time~\cite{chan1999more}.

For $k$-center with outliers, Charikar~\etal~\cite{charikar2001algorithms} 
gave the first algorithm with an approximation factor of $3$, 
that works in any dimension.
%which is improvable to 3, if all candidate center points can be enumerated.
%depending on whether the candidate center points can be enumerated or not, respectively.
Better results are known for small $k$ in the plane.
The 1-center problem with $z$ outliers in the plane can be solved 
in $\O(n \log n + z^{3} n^{\eps})$ time, for any $\eps > 0$, 
using Matou{\v{s}}ek's framework~\cite{matouvsek1995geometric}.
Agarwal~\cite{agarwal2008efficient} gave a randomized 
$\O(n z^{7} \log^3 z)$-time algorithm for 2-center with $z$ outliers in the plane.

\begin{figure}[t]
	\centering
	\includegraphics[width=12em]{figs/definition}
	\caption{$2$-center problem with 6 outlier}
	\label{fig:definition}
\end{figure}

In the streaming model, McCutchen~\etal~\cite{mccutchen2008streaming} 
and Guha~\cite{guha2009tight} presented algorithms to maintain $(2+\eps)$-approximation 
to the $k$-center problem in %in any metric space using 
$\O(\frac{kd}{\eps} \log \frac{1}{\eps})$ space.
For $k=1$, %better algorithms are known. % in high dimensions.
%a $1.5$-approximation in high dimensions was presented by Zarrabi-Zadeh and Chan~\cite{zarrabi2006simple}.
a factor-$((1+\sqrt{3})/{2})$ approximation 
was presented by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
in high dimensions, using $O(d)$ space. 
The approximation factor was later improved 
%to $(1+\sqrt{3})/{2}$ by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
to $1.22$ by Chan and Pathak~\cite{chan2014streaming}.
%If the dimension is fixed, a $(1 + \eps)$-approximation
%can be maintained in $\O({k}/{\eps^{((d-1)/2)}})$ space 
%using the notion of core-sets~\cite{zarrabi2008core}., 
%Ahn~\etal~\cite{ahn2014computing} gave a $(2+\eps)$-approximation algorithm using $\O(\frac{d}{\eps})$ space and update time, which was improved for Euclidean spaces by 
For $k=2$, Kim~\etal~\cite{kim2014improved} 
have recently obtained a $(1.8+\eps)$-approximation  
using $\O(\frac{d}{\eps})$ space and update time.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{\begin{tabular}[c]{@{}c@{}} \ \  Problem \ \ \end{tabular}} & \multicolumn{2}{c|}{Approximation Factor} \\ \cline{2-3} 
 & Without Outliers & With Outliers \\ \hline
1-Center & 1.22~\cite{chan2014streaming} & 1.73~\cite{zarrabi2009streaming} \\ \hline
2-Center & $1.8 + \eps$~\cite{kim2014improved} & \textbf{$1.8 + \eps$} [Here]  \\ \hline
$k$-Center & $2 + \eps$~\cite{guha2009tight,mccutchen2008streaming} & $4 + \eps$~\cite{mccutchen2008streaming}  \\ \hline
\end{tabular}
\caption {Summary of the streaming algorithms 
for $k$-center with and without outliers in high dimensions.}
\label{table:summary}
\end{table}

\paragraph{Our result}
In this paper, we study the 2-center problem with outliers in high dimensions.
We present a streaming algorithm for the problem that achieves
an approximation factor of $1.8 + \eps$, for any $\eps > 1$,
%using $O(d^3z/\eps)$ space.
using $\poly(d, z, {1 \over \eps})$ space and update time.
This improves the current streaming algorithm available for the problem
which has an approximation factor of $4+\eps$.
%Our work not only improves the best previous algorithm available for the case of 2-center,
%but is also counterintuitive in the sense that its 
The approximation factor of our algorithm matches
that of the best streaming algorithm for the 2-center problem without outliers.
This is somewhat surprising, considering that 
the best current approximation factors for streaming $k$-center with and without outliers 
differ by a factor of $\sqrt{2}$ for $k=1$,
and differ by a factor of 2  for general~$k$. %the 1-center problem.
See \Cref{table:summary} for a comparison.

To obtain our result, we have used a combination of several ideas
including parallelization construction, far/close ball separation,
and keeping a lower bound on the radius and distance of the optimal balls.
Our algorithm also utilizes the ideas used in~\cite{kim2014improved} 
for the 2-center problem with no outliers.
Nevertheless, our algorithm is much detailed and uses many new ingredients.
Indeed, handling outliers makes the $k$-center problem much harder,
as we not only need to find clusters with minimum radius, 
but we also need to decide which subset of points to cluster.
This is in particular more challenging in the streaming model,
where we only have a single pass over the input, and we must decide on the fly
which point can be an outlier, and which one can be safely dropped as a non-outlier point.

\REM{
The rest of this paper is organized as follows.
In Section~\ref{sec:1-center}, we give a simple $2$-approximation algorithm for 
the 1-center problem with outliers that uses $\O(z^2 + zd)$ space.
In the next section, we show a $(1.8 + \eps)$-approximation streaming algorithm for 2-center with $z$ outliers in high dimensions. 
%In Section~\ref{subsec:smaller} we run $z + 1$ parallel instances of the algorithm in order to find the first non-outlier point.
}

@@

\section{Preliminaries}
\label{sec:pre}
Let $B(c,r)$ denote a ball of radius $r$ centered at $c$. 
Given two points $p$ and $q$, we denote by $pq$ the straight line segment between $p$ and $q$,
and denote by $\len{pq}$ the length of this segment. 
Given two balls $B(c,r)$ and $B'(c',r')$, 
we define 
$\delta(B, B') = \max \set{0, \len{cc'}-r-r'}$
to be the \emph{distance} between $B$ and $B'$.
Two congruent balls $B_1$ and $B_2$ with radius $r$ are said to be \emph{$\alpha$-separated}, 
if $\delta(B_1, B_2) \geq \alpha r$.

@@

\begin{obs}
\label{obs:omitting-centerpoint}
	Given a set $P$ of $k(d+1)$ points in $d$-dimensional space, 
	there is a point $c$, not necessarily in $P$, 
	such that any convex object that does not cover $c$, 
	leaves at least $k$ points of $P$ uncovered.
\end{obs}

Given a point set $P$ and a point $p \in P$,
the \emph{$k$-furthest point from $p$}
is the $k$-th point of $P$ when points are sorted
in their decreasing distance from~$p$.

We assume the standard word-RAM model of computation. 
Each coordinate value takes a unit of space.
Thus, a point in $d$ dimensions takes $\O(d)$ space,
and basic operations on the points take $\O(d)$ time.

@@

\section{A 2-Approximation for 1-Center with Outliers}
\label{sec:1-center}

To warm up, we present a simple $2$-approximation streaming algorithm
for the 1-center problem with outliers.
It utilized a parallelization technique that 
will be used extensively in the rest of the paper.
The pseudocode is provided in \Cref{alg:1-center}.
The algorithm receives as input a stream of points, $P$,
the number of outliers, $z$, 
and a special point $c \in P$ which is guaranteed to be 
a non-outlier in the optimal solution.
It returns a ball $B$ covering all but at most $z$ points of $P$.
%We will show later how to remove this assumption.

@@

\begin{algorithm}
\caption{\sc 1-Center$(P, z, c)$} 
\algsetup{indent=1.5em}
\label{alg:1-center}
\begin{algorithmic}[1]
	%\STATE {$\triangleright$ $c$ is a non-outlier point in the optimal solution}
	\STATE $B \gets B(c, 0)$
	\STATE $Q \gets \emptyset$
	\FORALL {$p$ in $P$}
		\IF {$p \notin B$}
			\STATE insert $p$ into $Q$
			\IF {$|Q| = z + 1$}
				\STATE $q$ $\gets$ closest point to $c$ in $Q$
				\STATE remove $q$ from $Q$
				\STATE $B \gets B(c, \len{cq})$
			\ENDIF
		\ENDIF
	\ENDFOR
	\RETURN {B}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem} \label{thm:1-center}
	Algorithm~\ref{alg:1-center} computes a $2$-approximation to
	the 1-center problem with $z$ outliers,
	provided that a non-outlier point in the optimal solution in given.
\end{theorem}

\begin{proof}
Let $B^*(c^*, r^*)$ be the optimal solution,
and $c$ be a non-outlier in the optimal solution.
Since $c$ is covered by $B^*$,
for all points $p \in B^*$, we have
$\len{cp} \le \len{cc^*} + \len{c^*p} \le 2r^*$.
Among the $z+1$ points furthest from $c$,
there is at least one point $q$ which is not outlier in the optimal solution,
and therefore, is contained in $B^*$
(see \Cref{fig:1center}). 
Thus, $\len{cq} \le 2r^*$,
and therefore, the ball $B(c, \len{cq})$ returned by Algorithm~\ref{alg:1-center} 
is a 2-approximation. 
\end{proof}

\begin{figure}[t]
	\centering
	\includegraphics[width=12em]{figs/one-center}
	\caption{Proof of \Cref{thm:1-center}}
	\label{fig:1center}
\end{figure}

\noindent
Algorithm~\ref{alg:1-center} 
assumes that a point $c$ is known in advance
which is guaranteed to be a non-outlier in the optimal solution. 
To overcome the lack of knowledge about such a point, 
we run $z+1$ instances of \Cref{alg:1-center} in parallel,
each of which is given as $c$ one of the first $z + 1$ points in the stream.
Clearly, there exists a point among the first $z+1$ points of $P$
which is not an outlier in the optimal solution.
Therefore, the smallest ball among the $z+1$ balls computed in parallel
is always within factor~2 of the optimal solution. % by \Cref{thm:1-center-special}.
Since the space complexity of Algorithm~\ref{alg:1-center} for one instance is $\O(zd)$, 
and its update time is $\O (zd \log z)$, % and the query time is $\O (z)$.
we get the following result.

@@

Let $B_1^*$ and $B_2^*$ be the optimal balls
in an optimal solution to the 2-center problem with $z$ outliers on a point set $P$.
We denote by $r^*$ the optimal radius,
and by $\delta^*$ the distance between $B_1^*$ and $B_2^*$.
To prove our main result, we distinguish between two cases.
In \Cref{subsec:bigger}, we address the case where $\delta^* > \alpha r^*$, 
for some constant $\alpha$ to be fixed later.
(It will be turn out that $\alpha = 16$ is a proper choice.)
We then present in \Cref{subsec:smaller} 
our algorithm for the case of $\delta^* \leq \alpha r^*$.

@@

In this section, we present a $1.8$-approximation algorithm for the case 
where optimal balls are separated by a distance greater than $\alpha r^*$.
%The value of $\alpha$ will be fixed later in this section.

@@

\begin{obs}
\label{obs:intersection}
	Let $B_1$ and $B_2$ be two disjoint balls of distance equal to $\delta$,
	and let $B$ be an arbitrary ball of radius less than $\frac{\delta}{2}$. 
	Then $B$ intersects at most one of $B_1$ and $B_2$.
\end{obs}

@@

\begin{proof}
By Lemma~\ref{lem:c-sep}, there exists a point $q' \in B_2^*$ 
such that $\len{pq'} \ge \len{pq}$. 
Thus, by Observation~\ref{obs:c+4}, 
%$\frac{\alpha}{\alpha+4} \len{pp_2} \le \delta^*$. 
${\len{pq} \over \delta^*} \le {\len{pq'} \over \delta^*} < \frac{\alpha+4}{\alpha}$.
The claim follows immediately.
\end{proof}

\begin{lemma}
\label{lem:2r}
	If $p \in B_1^*(c_1, r^*)$ and $q \in B_2^*(c_2, r^*)$,
	then $B_1^* \subset B(p, 2r^*)$ and $B_2^* \subset B(q, 2r^*)$ and as a result at most $z$ points of $P$ lie outside $B(p, 2r^*) \cup B(q, 2r^*)$.
\end{lemma}

\begin{proof}
	For an arbitrary point $p' \in B_1^*$, $\len{pp'} \leq \len{pc_1} + \len{p'c_1} \leq 2r^*$ and as a result $B_1^* \subset B(p, 2r^*)$(With similar proof, $B_2^* \subset B(q, 2r^*)$).  Considering that at most $z$ points of $P$ is outlier,  the proof is complete.
\end{proof}

\begin{lemma}
\label{lem:center-point}
	Let $S$ be a subset of $P$ of size at least $\dz$,
	enclosed by a ball $B$ of radius less than $\delta^* / 2$.
	Then the centerpoint $c_p$ of $S$ lies either inside $B_1^{*}$ or inside $B_2^{*}$.
\end{lemma}

\begin{proof}
	By Observation~\ref{obs:intersection}, ball $B$ intersects exactly one of $B_1^*$ and $B_2^*$ (all points of $S$ can not be outlier). Without loss of generality, Let $B$ intersect with $B_1^*$. By Observation \ref{obs:omitting-centerpoint}, $c_p$ would lie in $B_1^*$ and the proof is complete.
\end{proof}

\paragraph{The Algorithm}
We now describe our algorithm for handling the case  $\delta^* > \alpha r^*$.
At any time, our algorithm maintains a partition of $P$ into three disjoint subsets
$B_1$, $B_2$, and Buffer.
The first point $p_1$ is assumed, w.l.o.g, to be in $B_1^*$. 
The algorithm tries to partition points in such a way that 
at the end,
$B_1$ contains the whole $B_1^*$, and $B_2$ contains the whole $B_2^*$,
with possibly some outliers being contained in $B_1$ and $B_2$.
%Buffer has always size at most $z$.
The algorithm sets $c_1 = p_1$ as the fixed center of $B_1$,
and picks $c_2$ among the points processed so far as a candidate center for $B_2$.
Moreover, the algorithm maintains two values $\delta$ and $r$,
where at any time, $\delta$ is a lower bound of $\delta^*$, 
and $r$ is an upper bound of $2r^*$ (if $c_p \in B_2^*$).

Our algorithm is presented in \Cref{alg:case-1}.
For each input point $p \in P$, the algorithm first tries 
to add $p$ to either $B_1$ or $B_2$,
using functions \textproc{AddToB$_1$} and \textproc{AddToB$_2$}, respectively.
%presented in Algorithms~\ref{alg:addB1} and~\ref{alg:addB2}.
If none of them fits, the point is added to Buffer.
The function \textproc{AddToB$_1$} adds a point $p$ to $B_1$
only if it is within distance $\delta$ of the center $c_1$.
Similarly, \textproc{AddToB$_2$} adds a point $p$ to $B_2$
only if it is within $\rc$-radius of $c_2$. 
The two functions also update $\delta$ and $\rc$ whenever necessary,
to maintain the invariants to be defined shortly in Lemma~\ref{lem:invariants}.

>>>>

%------------------------ Cleveref Labels -----------------------------

\Crefname{table}{Table}{Tables}
\Crefname{obs}{Observation}{Observations}
\Crefname{cor}{Corollary}{Corollaries}
\Crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{lemma}{Lemma}{Lemmas}

@@

\newcommand{\textproc}{\textsc}
\newcommand{\Call}[2]{\textsc{#1}(#2)}
\renewcommand{\algorithmicforall}{{\bf for each}}
\renewcommand{\algorithmiccomment}[1]{\quad $\vartriangleright$ #1}

@@

\begin{abstract}
We study the Euclidean 2-center problem with outliers in the data stream model. 
Given a stream of points in arbitrary $d$ dimensions, the goal is to find two congruent balls 
of minimum radius covering all but $z$ points. 
We provide a $(1.8+\eps)$-approximation streaming algorithm for the problem, 
improving upon the previous $(4 + \eps)$-approximation algorithm available for the problem.
The space complexity and update time of our algorithm is $\poly(d, z, {1 \over \eps})$,
independent of the size of the stream.
\end{abstract}

@@

\section{Introduction}
The $k$-center problem---covering a set of points 
using $k$ congruent balls of minimum radius---is a fundamental problem,
arising in many applications 
such as data mining, machine learning, statistics, and image processing.
In real-world applications, where input data is often noisy, 
it is very important to consider outliers, 
as even a small number of outliers can greatly affect the quality of the solution.
The $k$-center problem is particularly very sensitive to outliers,
and even a constant number of outliers can increase the radius of the $k$-center unboundedly.
Therefore, it is natural to consider the following generalization of
the the $k$-center problem: % which we call \emph{$k$-center with $z$ outliers}:
given a set $P$ of $n$ points in arbitrary $d$ dimensions
and a bound $z$ on the number of outliers,
find $k$ congruent balls of minimum radius 
to cover at least $n - z$ points of $P$.
See \Cref{fig:definition} for an example.
%for $z$ being the number of outliers.
%One of $k$-center problem variants is $k$-center with $z$-outlier which finds $k$ congruent balls with minimum radius covering all but $z$ points of a point set.

In this paper, we focus on the \emph{data stream} model of computation
in which only a single pass over input is allowed,
and we have only a limited amount of working space available.
%which is typically sublinear in the input size.
This model is specially useful for processing large data sets,
as it does not require the entire data set to be stored in memory.

@@

The Euclidean $k$-center problem has been extensively studied in the literature.
%There is a considerable amount of work on the geometric $k$-center problem.
If $k$ is part of the input,
the problem in known to be NP-hard in two and more dimensions~\cite{fowler1981optimal},
and is even hard to approximate to within a factor close to 2, 
unless P$\,=\,$NP~\cite{feder1988optimal}.
Factor-2 approximation algorithms are available for the problem 
in any dimension~\cite{gonzalez1985clustering,feder1988optimal}.
%The $k$-center problem without outliers (i.e., $z=0$) in general (Euclidean) space can not be approximated with a smaller factor than $2$ ($1.822$), unless $\mathbf{P}=\mathbf{NP}$~\cite{bern1996approximation}.
%Gonzalez~\cite{gonzalez1985clustering} showed a $2$-approximation algorithm for general case of the problem.
For small $k$ and $d$, %there are more efficient algorithms.
better solutions are available.
The 1-center problem in fixed dimensions is known to be LP-type 
and can be solved in $O(n)$ time~\cite{chazelle1996linear}.
For 2-center in the plane, the current best algorithm
runs in $\O(n \log^2 n \log^2 \log n)$ time~\cite{chan1999more}.

For $k$-center with outliers, Charikar~\etal~\cite{charikar2001algorithms} 
gave the first algorithm with an approximation factor of $3$, 
which works in any dimension.
%which is improvable to 3, if all candidate center points can be enumerated.
%depending on whether the candidate center points can be enumerated or not, respectively.
Better results are known for small $k$ in the plane.
The 1-center problem with $z$ outliers in the plane can be solved 
in $\O(n \log n + z^{3} n^{\eps})$ time, for any $\eps > 0$, 
using Matou{\v{s}}ek's framework~\cite{matouvsek1995geometric}.
Agarwal~\cite{agarwal2008efficient} gave a randomized 
$\O(n z^{7} \log^3 z)$-time algorithm for 2-center with $z$ outliers in the plane.

\begin{figure}[t]
	\centering
	\includegraphics[width=13em]{figs/definition}
	\caption{An example of $2$-center with 6 outliers}
	\label{fig:definition}
\end{figure}

In the streaming model, McCutchen~\etal~\cite{mccutchen2008streaming} 
and Guha~\cite{guha2009tight} presented algorithms to maintain $(2+\eps)$-approximation 
to the $k$-center problem in %in any metric space using 
$\O(\frac{kd}{\eps} \log \frac{1}{\eps})$ space.
For $k=1$, %better algorithms are known. % in high dimensions.
%a $1.5$-approximation in high dimensions was presented by Zarrabi-Zadeh and Chan~\cite{zarrabi2006simple}.
a factor-$((1+\sqrt{3})/{2})$ approximation 
was presented by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
in high dimensions, using $O(d)$ space. 
The approximation factor was later improved 
%to $(1+\sqrt{3})/{2}$ by Agarwal and Sharathkumar~\cite{agarwal2010streaming} 
to $1.22$ by Chan and Pathak~\cite{chan2014streaming}.
%If the dimension is fixed, a $(1 + \eps)$-approximation
%can be maintained in $\O({k}/{\eps^{((d-1)/2)}})$ space 
%using the notion of core-sets~\cite{zarrabi2008core}., 
%Ahn~\etal~\cite{ahn2014computing} gave a $(2+\eps)$-approximation algorithm using $\O(\frac{d}{\eps})$ space and update time, which was improved for Euclidean spaces by 
For $k=2$, Kim and Ahn~\cite{kim2014improved} 
have recently obtained a $(1.8+\eps)$-approximation  
using $\O(\frac{d}{\eps})$ space and update time.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{\begin{tabular}[c]{@{}c@{}} \ \  Problem \ \ \end{tabular}} & \multicolumn{2}{c|}{Approximation Factor} \\ \cline{2-3} 
 & Without Outliers & With Outliers \\ \hline \hline
1-Center & 1.22~\cite{chan2014streaming} & 1.73~\cite{zarrabi2009streaming} \\ \hline
2-Center & $1.8 + \eps$~\cite{kim2014improved} & \textbf{$1.8 + \eps$} [Here]  \\ \hline
$k$-Center & $2 + \eps$~\cite{guha2009tight,mccutchen2008streaming} & $4 + \eps$~\cite{mccutchen2008streaming}  \\ \hline
\end{tabular}
\caption {Summary of the streaming algorithms 
for $k$-center with and without outliers in high dimensions.}
\label{table:summary}
\end{table}

\paragraph{Our result}
In this paper, we study the 2-center problem with outliers in high dimensions.
We present a streaming algorithm for the problem that achieves
an approximation factor of $1.8 + \eps$, for any $\eps > 1$,
%using $O(d^3z/\eps)$ space.
using $\poly(d, z, {1 \over \eps})$ space and update time.
This improves the current streaming algorithm available for the problem
which has an approximation factor of $4+\eps$.
%Our work not only improves the best previous algorithm available for the case of 2-center,
%but is also counterintuitive in the sense that its 
The approximation factor of our algorithm matches
that of the best streaming algorithm for the 2-center problem without outliers.
This is somewhat surprising, considering that 
the best current approximation factors for streaming $k$-center with and without outliers 
differ by a factor of $\sqrt{2}$ for $k=1$,
and by a factor of 2  for general~$k$. %the 1-center problem.
See \Cref{table:summary} for a comparison.

To obtain our result, we have used a combination of several ideas
including parallelization construction, far/close ball separation,
and keeping a lower bound on the radius and distance of the optimal balls.
We have also utilized ideas used in~\cite{kim2014improved} 
for the 2-center problem with no outliers.
%However, our algorithm is much detailed and uses many new ingredients.
However, our problem is much harder here, % handling outliers makes the problem much harder,
as we not only need to find balls of minimum radius, 
but we also need to decide which subset of points to cluster.
This is in particular more challenging in the streaming model,
where we only have a single pass over the input, and we must decide on the fly
which point is an outlier, 
and which one can be safely ignored as a non-outlier point,
%using only a limited amount of storage.
to comply with the working space restriction enforced by the model.

\REM{
The rest of this paper is organized as follows.
In Section~\ref{sec:1-center}, we give a simple $2$-approximation algorithm for 
the 1-center problem with outliers that uses $\O(z^2 + zd)$ space.
In the next section, we show a $(1.8 + \eps)$-approximation
streaming algorithm for 2-center with $z$ outliers in high dimensions. 
}

@@

\section{Preliminaries}
\label{sec:pre}
Let $B(c,r)$ denote a ball of radius $r$ centered at $c$. 
We use $r(B)$ to denote the radius of a ball $B$. % and $c(B)$, respectively.
For two points $p$ and $q$, %by $pq$ the straight line segment between $p$ and $q$,
the distance between $p$ and $q$ is denoted by $\len{pq}$.
Given two balls $B(c,r)$ and $B'(c',r')$, 
we define 
$\delta(B, B') = \max \set{0, \len{cc'}-r-r'}$
to be the \emph{distance} between $B$ and $B'$.
Two balls $B_1$ and $B_2$ %with radius $r$ 
are said to be \emph{$\alpha$-separated}, 
%if $\delta(B_1, B_2) \geq \alpha r$.
if $\delta(B_1, B_2) \geq \alpha \cdot \max \set{r(B_1), r(B_2)}$.

@@

\begin{obs}
\label{obs:omitting-centerpoint}
	Given a set $P$ of $k(d+1)$ points in $d$-dimensional space, 
	the centerpoint of $P$ has the property that 
	any convex object not covering the centerpoint, 
	leaves at least $k$ points of $P$ uncovered.
\end{obs}

Given  a point set $P$,
the \emph{$k$-furthest point from} $p \in P$
is a point whose distance to $p$ is the $k$-th largest among all points in $P$.
%is the $k$-th point in $P$ when points are sorted
%in their decreasing distance from~$p$.
We assume the standard word-RAM model of computation. 
Each coordinate value takes a unit of space.
Thus, a $d$-dimensional point takes $\O(d)$ space,
and basic operations on the points take $\O(d)$ time.

@@

\section{A Simple Algorithm for 1-Center with Outliers}
\label{sec:1-center}

To warm up, we present a simple $2$-approximation streaming algorithm
for the 1-center problem with outliers.
It utilized a parallelization technique that 
will be used extensively in the rest of the paper.
The pseudocode is provided in \Cref{alg:1-center}.
The algorithm receives as input a stream of points, $P$,
and the number of outliers, $z$.
It also assumes that the first point $p_1$ of the stream is non-outlier.
We will show later how to remove this assumption.
The algorithm returns a ball $B$ covering all but at most $z$ points of $P$.
%

@@

\begin{algorithm}
\caption{\sc 1-Center$(P, z)$} 
\algsetup{indent=1.5em}
\label{alg:1-center}
\begin{algorithmic}[1]
	%\STATE {$\triangleright$ $c$ is a non-outlier point in the optimal solution}
	\STATE $B \gets B(p_1, 0)$
	\STATE $Q \gets \emptyset$
	\FORALL {$p$ in $P$}
		\IF {$p \notin B$}
			\STATE insert $p$ into $Q$
			\IF {$|Q| = z + 1$}
				\STATE $q$ $\gets$ closest point to $c$ in $Q$
				\STATE remove $q$ from $Q$
				\STATE $B \gets B(c, \len{cq})$
			\ENDIF
		\ENDIF
	\ENDFOR
	\RETURN {B}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem} \label{thm:1-center}
	Algorithm~\ref{alg:1-center} computes a $2$-approximation to
	the 1-center problem with $z$ outliers,
	assuming that the first point of the stream is not outlier. 
\end{theorem}

\begin{proof}
Let $B^*(c^*, r^*)$ be the optimal solution,
and $c = p_1$ be a non-outlier point in the optimal solution.
Since $c$ is covered by $B^*$,
for all points $p \in B^*$, we have
$\len{cp} \le \len{cc^*} + \len{c^*p} \le 2r^*$.
Among the $z+1$ points furthest from $c$,
there is at least one point $q$ which is not outlier, % in the optimal solution,
and therefore, is contained in $B^*$
(see \Cref{fig:1center}). 
Thus, $\len{cq} \le 2r^*$,
and hence, the ball $B(c, \len{cq})$ returned by Algorithm~\ref{alg:1-center} 
is a 2-approximation. 
\end{proof}

\begin{figure}[t]
	\centering
	\includegraphics[width=10em]{figs/one-center}
	\caption{Proof of \Cref{thm:1-center}}
	\label{fig:1center}
\end{figure}

\noindent
Algorithm~\ref{alg:1-center} 
assumes that the first point of the stream is not outlier.
To remove this assumption,
we run $z+1$ instances of \Cref{alg:1-center} in parallel,
each of which is given as input one of the first $z + 1$ points of the stream,
followed by the rest of the points.
Clearly, there exists a point among the first $z+1$ points of $P$
which is not an outlier in the optimal solution.
Therefore, the smallest ball among the $z+1$ balls computed in parallel
is always within factor~2 of the optimal solution. % by \Cref{thm:1-center-special}.
Since the space complexity of Algorithm~\ref{alg:1-center} for one instance is $\O(zd)$, 
and its update time is $\O (zd \log z)$, % and the query time is $\O (z)$.
we get the following result.

@@

Let $B_1^*$ and $B_2^*$ be the balls
in an optimal solution to the 2-center problem with $z$ outliers on a point set $P$.
We denote by $r^*$ the optimal radius,
and by $\delta^*$ the distance between $B_1^*$ and $B_2^*$.
To prove our main result, we distinguish between two cases.
In \Cref{subsec:bigger}, we address the case where $\delta^* > \alpha r^*$, 
for some constant $\alpha$ to be fixed later.
(It will turn out that $\alpha = 16$ is a proper choice.)
We then present in \Cref{subsec:smaller} 
our algorithm for the case of $\delta^* \leq \alpha r^*$.

@@

In this section, we present a $1.8$-approximation algorithm for the case 
where optimal balls are separated by a distance greater than $\alpha r^*$.
We start by two useful observations.
%The value of $\alpha$ will be fixed later in this section.

@@

\begin{obs}
\label{obs:intersection}
	Let $B_1$ and $B_2$ be two disjoint balls of distance $\delta$,
	and let $B$ be an arbitrary ball of radius less than $\frac{\delta}{2}$. 
	Then $B$ intersects at most one of $B_1$ and $B_2$.
\end{obs}

@@

\begin{proof}
By Lemma~\ref{lem:c-sep}, there exists a point $q' \in B_2^*$ 
such that $\len{pq'} \ge \len{pq}$. 
Thus, by Observation~\ref{obs:c+4}, 
%$\frac{\alpha}{\alpha+4} \len{pp_2} \le \delta^*$. 
${\len{pq} \over \delta^*} \le {\len{pq'} \over \delta^*} < \frac{\alpha+4}{\alpha}$.
\end{proof}

\begin{lemma}
\label{lem:2r}
	If $p \in B_1^*(c_1, r^*)$ and $q \in B_2^*(c_2, r^*)$,
	then $B_1^* \subset B(p, 2r^*)$ and $B_2^* \subset B(q, 2r^*)$,
	and hence, at most $z$ points of $P$ lie outside $B(p, 2r^*) \cup B(q, 2r^*)$.
\end{lemma}

\begin{proof}
	For an arbitrary point $p' \in B_1^*$, $\len{pp'} \leq \len{pc_1} + \len{p'c_1} \leq 2r^*$,
	and as a result, $B_1^* \subset B(p, 2r^*)$.
	Similarly, we have $B_2^* \subset B(q, 2r^*)$. 
	Considering that at most $z$ points of $P$ are outlier, the proof is complete.
\end{proof}

\begin{lemma}
\label{lem:center-point}
	Let $S$ be a subset of $P$ of size at least $\dz$,
	enclosed by a ball $B$ of radius less than $\delta^* / 2$.
	Then the centerpoint $c_p$ of $S$ lies inside either $B_1^{*}$ or $B_2^{*}$.
\end{lemma}

\begin{proof}
	Not all points in $S$ can be outlier, because $\dz > z$.
	Thus, by Observation~\ref{obs:intersection},
	$B$ intersects exactly one of $B_1^*$ and $B_2^*$.
	Assume, w.l.o.g., that $B$ intersect $B_1^*$. 
	Now, by Observation \ref{obs:omitting-centerpoint}, 
	if $c_p$ is not in $B_1^*$, 
	then $z+1$ points of $S$ remain uncovered by $B_1^*$,
	contradicting the fact that there at most $z$ outliers.	
\end{proof}

\paragraph{The Algorithm}
We now describe our algorithm for handling the case  $\delta^* > \alpha r^*$.
At any time, our algorithm maintains a partition of $P$ into three disjoint subsets
$B_1$, $B_2$, and Buffer.
The first point $p_1$ is assumed, w.l.o.g, to be in $B_1^*$. 
(We have already assumed that $p_1$ is not outlier.)
The algorithm tries to partition points in such a way that 
at the end,
$B_1$ contains the whole $B_1^*$, and $B_2$ contains the whole $B_2^*$,
with possibly some outliers being contained in $B_1$ and $B_2$.
%Buffer has always size at most $z$.
The algorithm sets $c_1 = p_1$ as the fixed center of $B_1$,
and picks $c_2$ among the points processed so far as a candidate 
for being the center of $B_2$.
Moreover, the algorithm maintains two values $\delta$ and $r$,
where at any time, $\delta$ is a lower bound of $\delta^*$, 
and $r$ is an upper bound of $2r^*$ (under a specific condition). %(if $c_p \in B_2^*$).
